Moisture absorption and drying alters nonadditive litter flammability in mixed conifer forest

### 2019-12-19 ms/response_to_reviews.tex

\name{Dylan W. Schwilk}

\begin{document}
\begin{letter}{}

\opening{Dear Dr. David Gibson,}

Enclosed is our revised manuscript research article, ``Moisture effect on leaf
litter alters non-additive flammability in mixed conifer forest''.

@@

We thank the editor and reviewer for their helpful comments and suggestions. We
have itemized each comment below and followed each with our reply and a list of
changes we made in response to the comment.

{\bf Responses to Senior Editor's comments:}

\begin{quote}
  I support the AE's recommendation to reject but allow resubmission. However,
  the bar for a resubmission will be high. My feeling is that your study
  confirms the essential findings from your 2012 study published in Journal of
  Ecology with the addition that your methodology of drying down adds some
  realism. Indeed, the confirmatory nature of this study was pointed out by
  reviewer 2 in their point 6. In addition to the points raised by the AE and
  reviewers, I will be looking in any resubmission for more evidence that this
  study advances your earlier work (apart from your last paragraph, your
  discussion is largely an explanation of your results with little context from
  other studies), that the findings have real ecological relevance (i.e., does
  it matter to the mixed-conifer communities?) and how is the new dry down
  treatment related to dry down conditions in litter mixtures in the field. If
  you can adequately address all of these concerns then I'll be happy to
  consider a resubmission.
\end{quote}

We agree that our prior Discussion was weak and have extensively rewritten it.
We argue that this work is not merely confirmatory. Our revised Introduction
argues for the ecological significance of dry down rates. The importance of
fuel moisture is well established and fire in the western US is often limited
by fuel moisture (Parks et al 2014). Furthermore, fine scale heterogeneity in
fire behavior has been statistically associated with variation in fuel
moisture (Hille and Stephens 2005) and fire history can vary over short spatial
scales as a consequence of species composition (Stephens 2001). Although the
importance of dry-down rates is less well-established, prescribed burns in
Sierra Nevada mixed conifer forest demonstrate that fuel moisture likely drives
differences in fire behavior depending on the timing of fire throughout the
season (Kaufman and Martin 1989, Knapp and Keeley 2006, Hille and Stephens
2005, Knapp et al 2006). These differences in fire behavior in turn control
tree mortality (Stephens et al 2002, Schwilk et al 2006). Despite this, we know
of only one other study specifically investigating differential drying behavior
(Kreye et al 2013).

We did not conduct a field experiment to compare drying rates in nature, but we
expect our experimental design to accelerate drying and explain that now in the
text. In terms of comparing fuel moisture with field measurements, Fuels can
vary from single digits to over 100\% fuel moisture throughout the fire season.
For example, 10-hour fuels in northern California varied from less than 10\% to
over 80\% depending on precipitation events in a study following fuel moisture
throughout the fire season (Estes et al 2012). Most measurements come from
prescribed burns where prescription guidelines mean that fires are rarely
attempted when 1 and 10 hour fuels are greater than 15\% moisture. For example,
early season burns in this mixed conifer system show evidence of moisture
limiting fire behavior at 12-14\% fuel moisture (Knapp et al 2005).

@@

Thank you. Several comments converged on the absolutely correct criticism that our Discussion did not engage adequately with the literature. We have made substantial changes to the Introduction and the Discussion adding more context and citing more of the recent literature. We reorganized the Introduction and now do not rush the reader so quickly through the ideas. We have expanded our justification for the study as argued for by the Senior Editor.

\begin{quote}
The authors mention leaf size explicitly as an important parameter in non-additivity (with some references including their own 2012 paper) but they do not include it in the current study. Especially at low moisture, the small individual needles of Abies should cause low ventilation and a high likelihood of negative non-additivity on flammability, as was shown by Zhao et al 2016 Ecol. Evol. Figure 4). In the Methods, as observed also by reviewer 2, there is very little detail on the extent to which litter of the respective species was picked up as individual leaves, branchlets without leaves versus branchlets with leaves still attached. This is really important, also because positive versus negative non-additivity of leaf-twig mixtures have been reported recently (Zhao et al. 2019 Plant and Soil). I feel that data and/or photos indicating the size and shape of the different litter types (and if possible variability in these) are important to add and to be used in the interpretation of the results.
\end{quote}

We have added information to the methods explaining the litter collection in
more detail. All material was naturally dropped litter which could include fine
twigs and leaves. Litter in these two \emph{Abies} species is a combination of
branchlets with still attached needles and individual needles that have
detached from the branchlet. We now cite the Zhao et al (2019) results when
discussing the mixture results.

When deciding how to limit the number of tests at the comparitive scale
(species as data points), we decided to not include size, because we believed
its influence on flammability is primarily through litter packing and was
therefore captured by litterbed density. However, in response to the request to
show particle sizes, we have added a figure (SI fig. 1) to the SI materials
showing litter particle size (the containing prism volumes) distributions for
each of these species. These data were collected from the stored litter. We did
not add these data to the main figures as they are not part of any statistical
test or main conclusions, but we are happy to do so if the editor suggests.

@@

\begin{quote}
  In the Methods, it was unclear how the various treatments and replicates were
  grouped over time, given that temperature and humidity in the fire facility
  could vary. Were there blocks in time with one replicate for each of the
  treatments (in random burning order) within a block? Or was there a different
  statistical design in time?
\end{quote}

Guided by the drying rates measured in the initial trials, we staggered
starting saturation times across species. This was aimed at allowing a
completely blocked design. However, because some ignition attempts were
unsuccessful, some days lacked the full complement of species while others,
later in the process, sometimes had more than one replicate per species burned.
We've added this information to the text [Lines X-X].

\begin{quote}
  Line 196 (see also reviewer 2): how can we assume that the quality of the
  stored litter is the same as that used in the previous study 3 years earlier,
  i.e. can these two datasets be safely combined?
\end{quote}

We have added detail as requested by reviewer 2. See lines X-X. All litter was
initially oven dried in large paper bags. The litter was then stored in the lab
at low humidity for several weeks on open racks. The litter was then bagged and
sealed and stored within cardboard boxes in a lab storage area. We saw no signs
of mold nor ongoing decomposition. We minimized handling of the bags to avoid
mechanical damage and changes to litter particle size.

@@

{\bf Responses to the Reviewer 1's comments}

\begin{quote}
  This work compares mixtures of fuels, relative to their individual
  components, in terms of moisture retention and flammability. Moisture
  dynamics were also investigated in relation to properties of leaves of the
  species forming litter. This information is important in understanding
  flammability of surface fuels in the field.

  The study has been well thought through and well executed, and the ms is
  generally well written. Stylistically, the main problem is that commas are
  missing in many sentences. This might sound trivial, but will really help the
  reader.
\end{quote}

We thank the reviewer for their comments and apologize for our failure in
proofreading. The reviewer should not have had to conduct copy editing and we
sincerely apologize for that. We have taken heed of the reviewers' comments and
fixed all specific grammar errors and missing symbols. For brevity, we've
omitted responses to the very simple suggested grammar fixes and simply made
all requested changes. We respond to other points individually.

@@

\begin{quote}
  line 72 - the comment about natural lightning ignitions comes from nowhere,
  and is a bit of a jolt for the reader - rephrase slightly, e.g. many fires
  result from lightning ignitions, which are concurrent with ....
\end{quote}

This portion was poorly written and this was also mentioned by Reviewer 2. We
have rewritten this in response to reviewer 2's comment that most fires that
spread result from dry lightning strikes across the west. See our responses to
Reviewer 2, below.

@@

\begin{quote}
  line 307 and line 326 - a minor inconsistency is that you analyse present
  species groups, yet discuss the results in terms of individual species
\end{quote}

We've tried to increase consistency throughout.

\begin{quote}
  lines 348-349 - logically, this does not necessarily follow - factors such as
  large leaves or high SLA could independently affect both absorptive capacity
  and flammability - although fuel moisture does of course have a strong effect
\end{quote}

We think that our logic must not have been clear here as we don't see the
contradiction. HOwever, We have rewritten these sentences and we hope they are
no longer confusing. See lines X-X.

@@

\begin{quote}
  line 365 - wording is problematic - change to '...ignition is possible, it
  had faster spread rates ...'
\end{quote}

Agreed. Fixed.

\begin{quote}
  line 365 - The low flammability of litter from your Cupressaceae species is
  interesting - Scarff and Westoby (2006) also found this for the Australian
  Cupressaceae genus Callitris, as did Bowman et al. (2018) J. Ecol 106, 1010
  -1022
\end{quote}

We have added some better discussion of past results and included discussion of
these papers. We also now contrast our result with that of Cornwell et al
(2015) as suggested out by the Associate Editor. See lines XX-XX.

@@

Yes. That omission was an oversight. We have added citations here. See lines
XX-XX

{\bf Responses to the Reviewer 2's comments}

\begin{quote}

  My major concerns are:

  1. One motivation for the study is not valid. I do not completely agree with
  the authors that natural lightning ignitions are concurrent with
  precipitation and, therefore, fire behavior can depend on rate of drying
  following a wetting event. The efficiency of individual lighting strikes in
  igniting a fire is affected by the moisture properties of forest fuel at the
  moment of lighting (Krawchuk et al. 2006, Ecology). Dry lighting that occurs
  without any accompanied rainfall nearby is most likely to cause wildfire
  because the fuel is dry enough to be ignited.
\end{quote}

Our sentence was poorly written and oversimplified. We thank the reviewer for
pointing this out and we have rewritten this section. However, our badly made
point was: lighting ignitions can occur during the same general period as
precipitation; therefore, drying rates are important. As the reviewer points
out, dry lightning strikes that find dry fuel are more likely to lead to fires
than are strikes that find damp fuels (Abatzoglou et al 2016). The fact that
the total number lightning strikes does not well predict burned area and that
ignitions occur often from dry strikes argues for the importance of fuel
moisture. Although synoptic weather conditions that create dry and wet storms
differ, the potential for precipitation and lightning during the same overall
period of thunderstorm activity often exists (eg wet and dry convection days
can occur under overlapping conditions, Rorig and Ferguson 1999). We've
rewritten this section and added details justifying the importance of drying
rates in fine fuels as suggested by the Senior Editor. See lines XX-XX.

@@

\begin{quote}
  2. The research question of the research is not explicitly stated both in the
  abstract and in the introduction. In the end of the introduction, the authors
  focus on specific data-collection goals, while leave the reader to figure out
  what the research question is.
\end{quote}

We have edited the abstract to make the research questions clearer. We agree
that our first submission jumped from stating current reserach gaps to specific
methods too quickly. See also our response to the Senior Editor, above.

@@

\begin{quote}
  3. The authors concluded that positive non-additivity appeared to be the rule
  in litter-driven fire and this non-additivity became more positive with fuel
  drying. But if we look closely at figure 8, non-additivity was negative
  within 50 hours since wetting, non-additivity was both positive and negative
  at about 75 hours since wetting, and non-additivity was positive at about 100
  hours since wetting. The high variability of non-additive effects at about 75
  hours since wetting might have important ecological effects on fire ignition
  and behavior, and are important information cannot be ignored.
\end{quote}

This is a good point and we have added some discussion of this. We tried to
limit our statements to those supported by the statistical tests (eg, the slope of
nonaddivity with drying was greater than 1), Therefore, we focused on the
positive effect of drying and the overall trend for positive nonadditivity.
This, of course, obscures some details which are worth mentioning and we have
edited this section in response to the reviewer's criticism. We agree that the
high variability in nonadditivity across mixtures deserves highlighting and
that is the reason we presented the actual predictions and departures from
prediction across time. Please see our changes on lines X-X.

@@

\begin{quote}
  4. The writing of this manuscript needs to be improved for clarity and
  accuracy. There are many typos in the text, and the using to scientific terms
  lack consistency through the text.
\end{quote}

We have made extensive edits in response to both reviewer's suggestions.
Furthermore, we have standardized our usage of some terms. We use the term
``maximum water absorption'' to indicate water content following saturation. Of
course, this is slightly inaccurate as the processes involved include both
absorption and adsorption. Our earlier draft often used the term ``maximum
water retention''. But we think this earlier wording was confusing as retention
implied and inverse to drying (and we were inconsistent in usage as the
reviewer points out).

@@

We have added one.

\begin{quote}
  Lines 50-61: in this paragraph, after the first sentence I think it is more
  logic to first explain the mechanism of moisture’s negative effects on fire,
  then explain the reason of the driving importance of dead fuel moisture
  content to surface fire ignition and behavior because fine dead fuels respond
  more readily to environmental changes than living fuels.
\end{quote}

We have rearranged the logic of the introduction as suggested.

\begin{quote}
Line 72: revise the description ``litter fuel driven system''.
\end{quote}

@@

We have rewritten this section and removed that phrase.

\begin{quote}
Line 75: “at any particular time”?
\end{quote}

We have rewritten this for clarity. See line XX.

\begin{quote}
Line 85: revise ``driven by by'' and ``emergent litter litter''.
\end{quote}

@@

Fixed.

\begin{quote}
Line 88: “similarly”?
\end{quote}

We were not sure what the reviewer's criticism was here but have rewritten this
section to be more detailed and precise.

@@

\begin{quote}
  Line 95: “absorption capacity” and “rate of moisture loss” are not correct
  descriptions of processes. Better say “absorption and retention”.
\end{quote}

Good point. We fixed that here and in another sentence elsewhere. Furthermore,
we've edited to explain what we mean by ``absorption'' because we are referring
to a litter bed and we include adsorption to litter surfaces in that
definition. We had originally tried to avoid the chemical terms because we are
observing a higher-level phenomenon but our previous language was confusing.

@@

\begin{quote}
  Line 361: revise “that that”.
\end{quote}

Fixed.

\begin{quote}
  Line 366: “but”?
\end{quote}

We've rewritten these sentences as ``Our two Cupressaceae, which have dense
litter and low maximum water absorption, had low flammability across moisture
levels.''

@@

\begin{quote}
  Line 401: “striking”？
\end{quote}

We removed the word.

\begin{quote}
  Line 410: “not met by the constituent species alone”?
\end{quote}

We don't understand the criticism. Perhaps the word ``alone'' is ambiguous? We
changed ``alone'' to ``in monoculture'' and hope that is clearer.

### 2019-12-20 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
fixup
Close to final R2?

\title{Moisture effect on leaf litter alters non-additive flammability in mixed conifer forest}

\running{Leaf litter flammability}

@@

\nwords{8200}
\ntables{4}
\nfig{8}
\nref{51}

@@

      \item Moisture dynamics varied across species with higher specific leaf
    area being associated with higher maximum absorption and lower litter bulk
    density being associated with faster drying rates. We examined two axes of
    flammability variation and found that these differing moisture dynamics
    result in time since wetting having strongly differing effects on
    flammability across the major litter types typified by different tree taxa.
    Litter mixtures exhibited nonadditivity in flame spread rate and
    in fuel consumption. This nonadditivity became more positive with fuel
    drying.

      \item \emph{Synthesis:} Taxon specific patterns of moisture absorption
    and retention can change flammability rankings under moist compared with
    dry fuels because traits that increase flame spread rate increase water
    absorption. Litter mixtures generally burn with fire behavior more similar
    to that of the most flammable constituent species and this effect increases
    as fuels dry.

@@

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In many forests that experience surface
fire, litter properties influence fire behavior
\citep{Ganteaume+Marielle+etal-2011, Schwilk+Caprio-2011,
  Varner+Kane+etal-2015}. In litter driven fire, the two major axes of
variation are described by measurements related to flame spread rate and
measurements related to heat release or heat transfer
\citep{Magalhaes+Schwilk-2012, Prior+Murphy+etal-2018}. These different
components of fire behavior have differing ecological effects with spread rate
being associated with rates of heat release and area burned, and with heat
transfer predicting soil impacts of fire and the mortality. A shortcoming in
our current knowledge is our integration of specific species and plant traits
into fuel hazard assessment, fire modeling, and fire effects prediction.
Developing quantitative yet robust and general ways to integrate actual plant
communities into fuel assessment will require understanding the magnitude of
flammability variability across species in space and time and parameterizing
based on plant characteristics.

@@

These species-specific flammability effects are increasingly well established.
However, how moisture dynamics modulate species specific flammability is poorly
known. Fuel moisture often limits fire activity in the western US
\citep{Miller+Urban-1999, Parks+Parisien-2014}. Fire scientists often
categorize fuel by size classes that correspond to moisture exchange rates,
although such simplified categories have long been recognized as not sufficient
to describe species-specific effects on moisture dynamics
\citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are dependent upon
forest structure (shading and rain throughfall effects), litter physical
structure, and chemical composition of the fuel-beds \citep{Nelson+Hiers-2008,
  Matthews-2014, Kreye_Hiers_etal-2018}. A variety of models describing
moisture dynamics of fine fuels are used by fire scientists \citep{Viney-1991,
  Nelson-2000, Catchpole+Catchpole+etal-2001}. These models usually aim to
describe equilibrium conditions and at least include temperature and relative
humidity although some include precipitation and solar radiation effects as
well. Explicit models of fire behavior as a function of time since wetting
could improve fire prediction, however. Drying rates following snowmelt or
following rainfall during the fire season will influence the length of the fire
season, flammability over time, and the probability of sustained ignition,
higher spread rates and fuel consumption when ignition occurs. That species can
differ in litter drying rates has been demonstrated for at least one forest
system \citep{Kreye+Varner+etal-2013}. Differential drying can be especially
important when precipitation can occur during the same season as ignitions.
Although synoptic weather conditions that create dry and wet lightning storms
differ, the potential for precipitation and lightning during the same overall
period of thunderstorm activity often exists: wet and dry convection days can
occur under overlapping conditions, \citep{Rorig+Ferguson-1999}. Moisture
levels of fine fuels respond rapidly to precipitation events during the fire
season \citep{Estes+Knapp+etal-2012}. Differential moisture content across time
may contribute to the often high spatial heterogeneity of surface fires
\citep{Hille+Stephens-2005, Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and to
variation in fire regime across landscapes \citep{Stephens-2001}.

Fuel moisture content through time is controlled by two processes, wetting
(including absorption and adsorption) and drying rate (desorbtion)
\citep{Anderson-1990, Kreye+Kobziar+etal-2013}. Absorptive capacity represents
the potential maximum fuel moisture and is evaluated by the moisture content of
litter when saturated; hereafter we refer to this measurement as ``maximum
water absorption''. Differences in moisture content over time since saturation
(drying rate) can indicate different abilities to retain moisture. Litter
particle geometry may have direct effects on maximum moisture absorption
through surface area affects on adsoprtion and we also expect effects of leaf
morphology, especially leaf cuticle thickness \citep{Van_Wagner-1969}. Drying
rate should also be influenced by leaf and litter bed traits:
surface-area-to-volume ratio and litter bulk density will influence air flow
through the litter bed. Litter packing, therefore, effects fire behavior in two
ways: directly by its effect on oxygen availability for combustion, and
indirectly by its effect on drying rate. Both these pathways are affected by
leaf particle traits, especially leaf size \citep{Scarff+Westoby-2006}.
Fuel-beds composed of larger leaves or leaves with lower maximum moisture
absorption should favor drying, thus increasing the probability of ignition and
fire spread.

@@

We asked if differences in litter moisture absorption and drying rates could
change the relative flammability of litter across species compared with
flammability measurements based on dry fuels. We investigated these litter
moisture dynamics across eight tree species in a temperate fire-prone forest in
California. We determined the maximum water absorption and rate of moisture
loss and examined how these traits vary in relation to specific leaf area and
litter packing density. Experimental burn trials of leaf litter at various
moisture levels determined how time since wetting influenced two axes of
flammability characterized by flame spread rate and total fuel consumed.
Previous work described the flammability of these species in monocultures and
in mixtures for oven-dry litter \citep{Magalhaes+Schwilk-2012} and found
consistent positive nonadditivity of flammability where fire behavior of
mixtures was driven by the most flammable species. We tested if the nonadditive
effects on reported in previous studies occur under higher moisture levels, and
whether fuel moisture effects strengthen or moderate any nonadditivity.

@@

We produced dry-down curves to determine maximum water absorption and the rate
of drying for each species and for four distinct three-species mixtures. These
curves allowed us to determine the rate at which a species loses moisture,
approximated as exponential decay, and to determine maximum water content
following saturation and gravity draining. Samples from the eight species (six
replicates per species) and four mixtures (5 replicates per mixture type) were
placed in baskets approximately 45 x 45\,cm by 15\,cm tall constructed of
aluminum mesh with 0.52\,cm square openings. Samples were 450\,g dry weight
litter per trial and litter depth was standardized to 10\,cm by changing the
horizontal dimensions of the basket. Weight was measured by using a balance
sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples
were immersed in plastic storage bins filled with water for 24 hours to fully
saturate them. After saturation, the baskets were removed from the water,
allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
content. The samples were then allowed to dry in a controlled environment kept
at $21^\circ$C temperature and 30\% relative humidity, assessed via an iButton
sensor (Maxim Integrated) placed over the samples. Litter baskets were arranged
in a completely randomized design. During the dry-down process three subsamples
of 5--10\,g were taken at 24 hour intervals and weighed. These subsamples were
then oven dried for 24 hours at $100^\circ$C and re-weighed to assess moisture
content. This was repeated until the final sample from a basket was less than
10\% moisture. This experimental setup was designed to accelerate drying
compared with field conditions because the baskets were suspended in air 2.8 --
5.3$^\circ$C above the warmest month average temperatures for the closest
weather stations to the collection sites (Grant Grove and Lodegpole stations,
NOAA 1981-2010 normals, https://www.ncdc.noaa.gov/cdo-web/datatools/normals).

With only eight species in this study, our power to test leaf trait effects on
the observed moisture dynamics was low, so we decided \emph{a priori} to test
the effect of one leaf trait likely to influence moisture absorption and
adsorption (specific leaf area), and one litter bed trait known to directly
affect flammability and likely to influence drying rate and litter packing
density (litter bulk density). Specific leaf area was measured on freshly
fallen litter by sampling 5 individuals per species. This measurement included
petiole and for the scale-leaved species included the full excised branchlet.
Litter bulk density was obtained in this experiment for each basket by dividing
the dry mass (450\,g) by the estimated litter volume (basket width x length x
litter depth). To examine litter particle size distributions in the stored
litter, we subsampled 10 approximately 2\,g samples per species from the stored
litter. Each individual particle of that subsample was measured for length
(longest axis), width (longest axis perpendicular to length) and thickness
(perpendicular to length and width at midrib) using an electronic caliper.

@@

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timing so as to attempt to burn samples over a range of moisture contents.
We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
preliminary trials on damp fuels suggested that across these fuels ignition
will not occur above 100\% fuel moisture. To expand our data on the dry end of
this gradient, we supplemented these data with data from an earlier experiment
on oven-dried litter \citep{Magalhaes+Schwilk-2012}. This supplemental data
added five replicates per species at an oven dried moisture content of $<$ 5\%.
We assigned these replicates a time since wetting value of 144 hours equivalent
to the longest time since drying for the wetted samples. Flammability was
assessed using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which
leaf litter was placed and the table gently shaken to allow settling. Burning
trials all used 450\,g (dry weight) of litter. For three-species mixtures, this
was 150\,g per species. Therefore, fuel mass and burning tray length and width
were fixed, but fuel depth was allowed to vary. We recorded the actual fuel
depth at three points along the tray to calculate litter bed density before
burning. This design allows fire to reach constant flame spread rates
\citep{Magalhaes+Schwilk-2012}. Two metal rulers equally spaced along the
apparatus allowed visual determination of maximum flame height. Litter was
ignited with a propane torch; maximum adiabatic flame temperature of
1899$^\circ$C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end
of the apparatus and allowed to burn until extinction. If a sample failed to
ignite after 30 seconds, no further attempt was made and if three samples of a
species at a particular moisture level failed to ignite, no further attempts
were made on that species/moisture combination and the replicate was allowed to
continue drying until the next scheduled checkpoint. The flammability
trials followed the methodology detailed in \citet{Magalhaes+Schwilk-2012}.

@@

Burn trials occurred in a cement structure used to simulate house fires at the
Fire Department of the City of Lubbock. This structure minimized wind and
helped regulate temperature and relative humidity. Temperature and relative
humidity were measured before every trial using a Kestrel 3000
(Nielsen-Kellerman, Boothwyn, PA). All trials were conducted on clear days
between 10 am and 3 pm. We scheduled burns in a randomized blocked design
aiming to burn one replicate of each species or mixture on single burn day.
This required scheduling species saturation and during start based on the dry
down curves.

@@

\subsection*{Species specific moisture dynamics and leaf traits}

>>>>

\title{Moisture absorption and drying alters nonadditive litter flammability
  in mixed conifer forest}

\running{Litter flammability}

@@

\nwords{8200}
\ntables{4}
\nfig{8}
\nref{55}

@@

      \item Moisture dynamics varied across species with higher specific leaf
    area being associated with higher maximum absorption and lower litter bulk
    density being associated with faster drying rates. We examined two axes of
    flammability variation and found that these differing moisture dynamics
    result in time since wetting having strongly differing effects on
    flammability across the major litter types typified by different tree taxa.

    \item Litter mixtures exhibited nonadditivity in flame spread rate and
    in fuel consumption. This nonadditivity became more positive with fuel
    drying.

      \item \emph{Synthesis:} Taxon specific patterns of moisture absorption
    and drying can change flammability rankings under moist compared with dry
    fuels because traits that increase flame spread rate increase water
    absorption. Litter mixtures generally burn with fire behavior more similar
    to that of the most flammable constituent species and this effect increases
    as fuels dry.

@@

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In many forests that experience surface
fire, litter properties influence fire behavior
\citep{Ganteaume+Marielle+etal-2011, Schwilk+Caprio-2011,
  Varner+Kane+etal-2015}. In litter driven fire, the two major axes of
variation are described by measurements related to flame spread rate and
measurements related to heat release or heat transfer
\citep{Magalhaes+Schwilk-2012, Prior+Murphy+etal-2018}. These different
components of fire behavior have differing ecological effects with spread rate
being associated with rates of heat release and area burned, and with heat
transfer predicting soil impacts of fire and tree mortality. A shortcoming in
our current knowledge is our integration of specific species and plant traits
into fuel hazard assessment, fire modeling, and fire effects prediction.
Developing quantitative yet robust and general ways to integrate actual plant
communities into fuel assessment will require understanding the magnitude of
flammability variability across species in space and time and parameterizing
based on plant characteristics.

@@

These species-specific flammability effects are increasingly well established.
However, how moisture dynamics modulate species specific flammability is poorly
known. Fuel moisture often limits fire activity in the western US
\citep{Miller+Urban-1999, Parks+Parisien-2014}. Fire scientists categorize fuel
by size classes that correspond to moisture exchange rates, although such
simplified categories have long been recognized as not sufficient to describe
species-specific effects on moisture dynamics \citep{Anderson-1985} which are
dependent upon forest structure (shading and rain through-fall effects), litter
physical structure, and chemical composition of the fuel-beds
\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity, although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however. Long term drying following snow
melt will influence the length of the fire season. Short term drying following
rainfall during the fire season will influence the probability of sustained
ignition, spread rates, and fuel consumption when ignition occurs. That species
can differ in litter drying rates has been demonstrated for at least one forest
system \citep{Kreye+Varner+etal-2013}.

Differential drying is especially important when precipitation occurs
during the same season as ignitions. Although synoptic weather conditions that
create dry and wet lightning storms differ, the potential for precipitation and
lightning during the same overall period of thunderstorm activity often exists:
wet and dry convection days can occur under overlapping conditions,
\citep{Rorig+Ferguson-1999}. Moisture levels of fine fuels respond rapidly to
precipitation events during the fire season \citep{Estes+Knapp+etal-2012}.
Differential moisture content across time may contribute to the often high
spatial heterogeneity of surface fires \citep{Hille+Stephens-2005,
  Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and to variation in fire regime
across landscapes \citep{Stephens-2001}.

Fuel moisture content through time is controlled by two processes, wetting
(including absorption and adsorption) and drying rate (desorbtion)
\citep{Anderson-1990, Kreye+Kobziar+etal-2013}. Absorptive capacity represents
the potential maximum fuel moisture and is evaluated by the moisture content of
litter when saturated; hereafter we refer to this measurement as ``maximum
water absorption''. Differences in moisture content over time since saturation
can indicate different abilities to retain moisture. Leaf morphology,
especially leaf cuticle thickness, can influence water absorption
\citep{Van_Wagner-1969}. Litter particle geometry may have effects on maximum
moisture content because a larger total surface area will increase water
adsorption. Drying rate should also be influenced by leaf and litter bed traits
because surface-area-to-volume ratio and litter bulk density will influence air
flow through the litter bed. Therefore, we expect litter packing to effect fire
behavior in two ways: directly by its effect on oxygen availability for
combustion, and indirectly by its effect on drying rate. Both these pathways
are affected by leaf particle traits, especially leaf size
\citep{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves or leaves with
thin cuticles should favor moisture absorption. On the other hand, large leaves
lower litter density and should favor drying, thus increasing the probability
of ignition and fire spread. The interaction of mositure absorption and drying
deserves investigation.

@@

We asked if differences in litter moisture absorption and drying rates could
change the relative flammability of litter across species compared with
flammability measurements based on dry fuels. We investigated these litter
moisture dynamics across eight tree species in a temperate fire-prone forest in
California. We determined the maximum water absorption and rate of moisture
loss and examined how these traits vary in relation to specific leaf area and
litter packing density. Experimental burn trials of leaf litter at various
moisture levels determined how time since wetting influenced two axes of
flammability characterized by flame spread rate and total fuel consumed. To
determine the potential magnitude of mixture effects, we tested if the
nonadditive effects on reported in previous studies occur under higher moisture
levels, and whether fuel moisture effects strengthen or moderate any
nonadditivity.

@@

We produced dry-down curves to determine maximum water absorption and the rate
of drying for each species and for four distinct three-species mixtures. These
curves allowed us to determine the rate at which a species loses moisture,
approximated as exponential decay, and to determine maximum water content
following saturation and gravity draining. Samples from the eight species (six
replicates per species) and four mixtures (5 replicates per mixture type) were
placed in baskets approximately 45 x 45\,cm by 15\,cm tall constructed of
aluminum mesh with 0.52\,cm square openings. Samples were 450\,g dry weight
litter per trial and litter depth was standardized to 10\,cm by changing the
horizontal dimensions of the basket. Weight was measured by using a balance
sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples
were immersed in plastic storage bins filled with water for 24 hours to fully
saturate them. After saturation, the baskets were removed from the water,
allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
content. The samples were then allowed to dry in a controlled environment kept
at $21^\circ$C temperature and 30\% relative humidity, assessed via an iButton
sensor (Maxim Integrated) placed 1\,m above the samples. Litter baskets were
arranged in a completely randomized design. During the dry-down process three
subsamples of 5--10\,g were taken at 24 hour intervals and weighed. These
subsamples were then oven dried for 24 hours at $100^\circ$C and re-weighed to
assess moisture content. This was repeated until the final sample from a basket
was less than 10\% moisture. This experimental setup was designed to accelerate
drying compared with field conditions because the baskets were suspended in air
2.8 -- 5.3$^\circ$C above the warmest month average temperatures for the
closest weather stations to the collection sites (Grant Grove and Lodegpole
stations, NOAA 1981-2010 normals,
https://www.ncdc.noaa.gov/cdo-web/datatools/normals).

With only eight species in this study, our power to test leaf trait effects on
the observed moisture dynamics was low, so we decided \emph{a priori} to test
the effect of one leaf trait likely to influence moisture absorption and
adsorption (specific leaf area), and one litter bed trait known to directly
affect flammability and likely to influence drying rate (litter bulk density).
Specific leaf area was measured on freshly fallen litter by sampling 5
individuals per species. This measurement included the petiole and for the
scale-leaved species included the full excised branchlet. Litter bulk density
was obtained in this experiment for each basket by dividing the dry mass
(450\,g) by the estimated litter volume (basket width x length x litter depth).
To examine litter particle size distributions in the stored litter, we
subsampled 10 approximately 2\,g samples per species from the stored litter.
Each individual particle of that subsample was measured for length (longest
axis), width (longest axis perpendicular to length) and thickness
(perpendicular to length and width at midrib) using an electronic caliper.

@@

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timing so as to attempt to burn samples over a range of moisture contents.
We staggered saturation and start of drying so we could conduct our burning in
a blocked design with multiple species at the same moisture content per burning
day. We aimed for moisture contents from 10\% to 100\% on a dry mass basis because
preliminary trials on damp fuels suggested that across these fuels ignition
will not occur above 100\% fuel moisture. To expand our data on the dry end of
this gradient, we supplemented these data with data from an earlier experiment
on oven-dried litter \citep{Magalhaes+Schwilk-2012}. This supplemental data
added five replicates per species at an oven dried moisture content of $<$ 5\%.
We assigned these replicates a time since wetting value of 144 hours equivalent
to the longest time since drying for the wetted samples.

Burn trials occurred in a cement structure used to simulate house fires at the
Fire Department of the City of Lubbock. This structure minimized wind and
helped regulate temperature and relative humidity. Temperature and relative
humidity were measured before every trial using a Kestrel 3000
(Nielsen-Kellerman, Boothwyn, PA) and these values were used to calculate vapor
pressure deficit as a single weather covariate at each trial. All trials were
conducted on clear days between 10 am and 3 pm. We scheduled burns in a
randomized blocked design aiming to burn one replicate of each species or
mixture on single burn day. This required scheduling species saturation and
during start based on the dry down curves.

Flammability was assessed using a 150\,cm long burn table, 15\,cm wide and
15\,cm tall, in which leaf litter was placed and the table gently shaken to
allow settling. Burning trials all used 450\,g (dry weight) of litter. For
three-species mixtures, this was 150\,g per species. Therefore, fuel mass and
burning tray length and width were fixed, but fuel depth was allowed to vary.
We recorded the actual fuel depth at three points along the tray to calculate
litter bed density before burning. This design allows fire to reach constant
flame spread rates \citep{Magalhaes+Schwilk-2012}. Two metal rulers equally
spaced along the apparatus allowed visual determination of maximum flame
height. Litter was ignited with a propane torch; maximum adiabatic flame
temperature of 1899$^\circ$C,
\url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the apparatus
and allowed to burn until extinction. If a sample failed to ignite after 30
seconds, no further attempt was made and if three samples of a species at a
particular moisture level failed to ignite, no further attempts were made on
that species/moisture combination and the replicate was allowed to continue
drying until the next scheduled checkpoint. The flammability trials followed
the methodology detailed in \citet{Magalhaes+Schwilk-2012}.

@@

@@

\subsection*{Moisture dynamics across species}
